Concurrent Programming with CXL

### 2020-06-26 book/paper.tex
Working on fonts.  Finished up to Chapter 9

CXL in fact supports this.
It has an operator \texttt{atLabel}~$L$,
\index{atLabel operator}
where $L$
is the atom containing the name of the label (in this case, \texttt{.cs}).
The operator returns a bag (multiset) of name tags of processes executing at that
label.  The bag is represented by a dictionary that maps each element
in the bag to the number of times the element appears in the bag.
Method \texttt{atLabel} only exists for specification purposes---do not
use it in normal code.
The assertion also makes use of the \texttt{nametag()} operator
\index{nametag operator}
that returns the name tag of the current process.
If you run the code through CXL, the assertion should fail because
there is no code yet for safely entering and exiting the critical section.

@@

While the CVM does not have any specific built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language in two ways.
First, any CXL statement can be made atomic by placing a label in front
of it.  Second, a group of CXL statements can be made atomic
through its \texttt{atomic}
\index{atomic statement}
statement.
We can use \texttt{atomic} statement blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\cxlsource{atomicinc}
\end{code}

Many CPUs have an atomic ``test-and-set'' (TAS)
\index{test-and-set}
\index{TAS}
operation.
Method \texttt{tas} in Figure~\ref{fig:tas} shows its specification.
Here \texttt{s} points to a shared boolean variable and \texttt{p}
to a private boolean variable, belonging to some process.
The operation copies the value of the shared variable to the
private variable (the ``test'')
and then sets the shared variable to \texttt{True} (``set'').

Figure~\ref{fig:tas} goes on how to implement mutual exclusion for
a set of $N$ processes.
The approach is called \emph{spinlock},
\index{spinlock}
because each process is ``spinning'' until
it can acquire the lock.
The program uses $N+1$ variables.
Variable \textit{shared} is initialized to \texttt{False} while
\textit{private}[$i$] for each process $i$ is initialized to \texttt{True}.
An important invariant, $\mathcal{I}_1$, of the program is that at any time at most
one of these variables is \texttt{False}.
Another invariant, $\mathcal{I}_2(i)$, is that if process $i$ is in the critical section,
then \textit{private}[$i$] == \texttt{False}.
Between the two (i.e., $\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$),
it is clear that only one process can be in the
critical section at the same time.

@@

We can do this by adding another process that, in a loop,
checks the invariant.  Figure~\ref{fig:tasinv} shows the code.
Method \texttt{checkInvariant()} checks to see if the invariant holds
in a state.  It introduces a new feature of CXL: the ability to have
variables local to a method.  In this case, the process variable \texttt{sum}
is used to compute the number of shared variables that have value
\texttt{False}.
The function is invoked by in a loop by a process that runs alongside
the other processes.
In CXL, \textbf{assert} statements are executed atomically, so the
evaluation of the assertion is not interleaved with the execution
of other processes.
Because CXL tries every possible execution, the process is guaranteed
to find violations of the invariant if it does not hold.

@@

In Figure~\ref{fig:tas} we have shown a solution based on a shared
variable and a private variable for each process.   The private
variables themselves are actually implemented as shared variables,
but they are accessed only by their respective processes.
There is no need to keep \texttt{private} as a shared
variable---we only did so to be able to show and check the invariants.
Figure~\ref{fig:tas2} shows a more straightforward implementation of spinlock.
The solution is similar to the na\"{i}ve solution of Figure~\ref{fig:uplock},
but uses test-and-set to check and set the lock variable atomically.
This approach is general for any number of processes.

It is important to appreciate the difference between an
\emph{atomic section} (the statements executed within an
\texttt{atomic} statement) and a \emph{critical section}
(protected by a lock of some sort).
The former ensures that while the
atomic statement is executing no other process can execute.
The latter allows multiple processes to run concurrently,
just not within the critical section.
The former is rarely available to a programmer, while the latter
is very common.

In CXL, atomic statements allow you to \emph{implement} your own
synchronization primitives like test-and-set.  Other common examples
include compare-and swap and fetch-and-add.  Atomic statements
are not intended to \emph{replace} locks or other synchonization primitives.
When solving synchronization problems you should not directly use
atomic statement but use the synchronization primitives that are available
to you.  But if you want to design a new synchronization primitive, then
use \texttt{atomic} by all means.

@@

\begin{code}
\begin{verbatim}
stop L
\end{verbatim}
\end{code}

Here \texttt{L} is a shared list.
Another process can revive the process using the \texttt{go}
\index{go statement}
statement:

\begin{code}
\begin{verbatim}
go C R
\end{verbatim}
\end{code}

Here \texttt{C} is a context and \texttt{R} is a CXL value.
It causes a process with context \texttt{C} to be added to the state that has
just executed the \texttt{stop}
\index{stop expression}
expression.  The \texttt{stop} expression returns the value \texttt{R}.

There is a second version of the \texttt{synch} module that uses suspension
instead of spinlocks.
Figure~\ref{fig:suspension} shows the same \texttt{Lock} interface implemented
using suspension.
The interface is implemented as follows:
\begin{itemize}
\item A \texttt{Lock} maintains both a boolean indicating whether the
lock is taken and a list of contexts of processes that want to acquire the lock.
\item
\texttt{lock}
\index{lock}
sets the lock if available and suspends itself if not.
Note that \texttt{stop} is called within an \texttt{atomic} statement---this is
the only exception to an atomic statement running to completion.  While the
process is running no other processes can run, but when it suspends itself
it allows other processes to run.
\item
\texttt{unlock}
\index{unlock}
checks to see if any processes are waiting to get the lock.
If so, it uses the \texttt{head} and \texttt{tail}
methods from the \texttt{list} module (see Appendix~\ref{ap:module})
to resume the first process that got
suspended and to remove its context from the list.
\end{itemize}
Selecting the first process is a design choice.  Another implementation could
have picked the last one, and yet another implementation could have used
\texttt{choose} to pick an arbitrary one.  Selecting the first is a common
choice in lock implementations as it prevents \emph{starvation}:
\index{starvation}
every process
gets a chance to obtain the lock (assuming every process eventually releases
it).  Chapter~\ref{ch:starvation} will talk more about starvation and how
to prevent it.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Boolean & \texttt{False}, \texttt{True} \\
\hline
integers & \texttt{-inf, ..., -2, -1, 0, 1, 2, ..., inf} \\
\hline
Atom & \texttt{.example}, \texttt{.test1} \\
\hline
Program Counter & (method names are program counter constants) \\
\hline
Dictionary & \texttt{dict\{ .account: 12345, .valid: False \} } \\
\hline
Set & \texttt{\{ 1, 2, 3 \}, \{ False, .id, 3 \} } \\
\hline
Address & \texttt{\&lock, \&flags[2], None} \\
\hline
Context & (generated by \texttt{stop} expression) \\
\hline
\end{tabular}
\vspace{1em}

>>>>

CXL in fact supports this.
It has an operator \textbf{atLabel}~$L$,
\index{atLabel operator}
where $L$
is the atom containing the name of the label (in this case, \texttt{.cs}).
The operator returns a bag (multiset) of name tags of processes executing at that
label.  The bag is represented by a dictionary that maps each element
in the bag to the number of times the element appears in the bag.
Method \textbf{atLabel} only exists for specification purposes---do not
use it in normal code.
The assertion also makes use of the \textbf{nametag}() operator
\index{nametag operator}
that returns the name tag of the current process.
If you run the code through CXL, the assertion should fail because
there is no code yet for safely entering and exiting the critical section.

@@

While the CVM does not have any specific built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language in two ways.
First, any CXL statement can be made atomic by placing a label in front
of it.  Second, a group of CXL statements can be made atomic
through its \textbf{atomic}
\index{atomic statement}
statement.
We can use \textbf{atomic} statement blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\cxlsource{atomicinc}
\end{code}

Many CPUs have an atomic ``test-and-set'' (TAS)
\index{test-and-set}
\index{TAS}
operation.
Method \texttt{tas} in Figure~\ref{fig:tas} shows its specification.
Here $s$ points to a shared boolean variable and $p$
to a private boolean variable, belonging to some process.
The operation copies the value of the shared variable to the
private variable (the ``test'')
and then sets the shared variable to \texttt{True} (``set'').

Figure~\ref{fig:tas} goes on how to implement mutual exclusion for
a set of \texttt{N} processes.
The approach is called \emph{spinlock},
\index{spinlock}
because each process is ``spinning'' until
it can acquire the lock.
The program uses $\mathtt{N}+1$ variables.
Variable \textit{shared} is initialized to \texttt{False} while
\textit{private}[$i$] for each process $i$ is initialized to \texttt{True}.
An important invariant, $\mathcal{I}_1$, of the program is that at any time at most
one of these variables is \texttt{False}.
Another invariant, $\mathcal{I}_2(i)$, is that if process $i$ is in the critical section,
then \textit{private}[$i$] == \texttt{False}.
Between the two (i.e., $\mathcal{I}_1 \land \forall i: \mathcal{I}_2(i)$),
it is clear that only one process can be in the
critical section at the same time.

@@

We can do this by adding another process that, in a loop,
checks the invariant.  Figure~\ref{fig:tasinv} shows the code.
Method \texttt{checkInvariant()} checks to see if the invariant holds
in a state.  It introduces a new feature of CXL: the ability to have
variables local to a method.  In this case, the process variable \textit{sum}
is used to compute the number of shared variables that have value
\texttt{False}.
The function is invoked by in a loop by a process that runs alongside
the other processes.
In CXL, \textbf{assert} statements are executed atomically, so the
evaluation of the assertion is not interleaved with the execution
of other processes.
Because CXL tries every possible execution, the process is guaranteed
to find violations of the invariant if it does not hold.

@@

In Figure~\ref{fig:tas} we have shown a solution based on a shared
variable and a private variable for each process.   The private
variables themselves are actually implemented as shared variables,
but they are accessed only by their respective processes.
There is no need to keep \textit{private} as a shared
variable---we only did so to be able to show and check the invariants.
Figure~\ref{fig:tas2} shows a more straightforward implementation of spinlock.
The solution is similar to the na\"{i}ve solution of Figure~\ref{fig:uplock},
but uses test-and-set to check and set the lock variable atomically.
This approach is general for any number of processes.

It is important to appreciate the difference between an
\emph{atomic section} (the statements executed within an
\textbf{atomic} statement) and a \emph{critical section}
(protected by a lock of some sort).
The former ensures that while the
atomic statement is executing no other process can execute.
The latter allows multiple processes to run concurrently,
just not within the critical section.
The former is rarely available to a programmer, while the latter
is very common.

In CXL, atomic statements allow you to \emph{implement} your own
synchronization primitives like test-and-set.  Other common examples
include compare-and swap and fetch-and-add.  Atomic statements
are not intended to \emph{replace} locks or other synchonization primitives.
When solving synchronization problems you should not directly use
atomic statement but use the synchronization primitives that are available
to you.  But if you want to design a new synchronization primitive, then
use \textbf{atomic} by all means.

@@

\begin{code}
\textbf{stop} $L$
\end{code}

Here $L$ is a shared variable containing a list.
Another process can revive the process using the \textbf{go}
\index{go statement}
statement:

\begin{code}
\textbf{go} $C$ $R$
\end{code}

Here $C$ is a context and $R$ is a CXL value.
It causes a process with context $C$ to be added to the state that has
just executed the \textbf{stop}
\index{stop expression}
expression.  The \textbf{stop} expression returns the value $R$.

There is a second version of the \texttt{synch} module that uses suspension
instead of spinlocks.
Figure~\ref{fig:suspension} shows the same \texttt{Lock} interface implemented
using suspension.
The interface is implemented as follows:
\begin{itemize}
\item A \texttt{Lock} maintains both a boolean indicating whether the
lock is taken and a list of contexts of processes that want to acquire the lock.
\item
\texttt{lock}
\index{lock}
sets the lock if available and suspends itself if not.
Note that \textbf{stop} is called within an \textbf{atomic} statement---this is
the only exception to an atomic statement running to completion.  While the
process is running no other processes can run, but when it suspends itself
it allows other processes to run.
\item
\texttt{unlock}
\index{unlock}
checks to see if any processes are waiting to get the lock.
If so, it uses the \texttt{head} and \texttt{tail}
methods from the \texttt{list} module (see Appendix~\ref{ap:module})
to resume the first process that got
suspended and to remove its context from the list.
\end{itemize}
Selecting the first process is a design choice.  Another implementation could
have picked the last one, and yet another implementation could have used
\texttt{choose} to pick an arbitrary one.  Selecting the first is a common
choice in lock implementations as it prevents \emph{starvation}:
\index{starvation}
every process
gets a chance to obtain the lock (assuming every process eventually releases
it).  Chapter~\ref{ch:starvation} will talk more about starvation and how
to prevent it.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Boolean & \texttt{False}, \texttt{True} \\
\hline
integers & \texttt{-inf, ..., -2, -1, 0, 1, 2, ..., inf} \\
\hline
Atom & \texttt{.example}, \texttt{.test1} \\
\hline
Program Counter & (method names are program counter constants) \\
\hline
Dictionary & \texttt{dict\{ .account: 12345, .valid: False \} } \\
\hline
Set & \texttt{\{ 1, 2, 3 \}, \{ False, .id, 3 \} } \\
\hline
Address & \texttt{\&lock, \&flags[2], None} \\
\hline
Context & (generated by \textbf{stop} expression) \\
\hline
\end{tabular}
\vspace{1em}
